Density-dependent selection in evolutionary genetics: a lottery model of Grime's triangle

### 2017-04-14 manuscript.tex
latest

Evolutionary models differ greatly in their treatment of fitness. In models of genetic evolution, genotypes are typically assigned constant (or frequency-dependent) selection coefficients describing the change in their relative frequencies over time due to differences in viability. This considerably simplifies the mathematics of selection, facilitating greater genetic realism, and can be justified over sufficiently short time intervals \citep[p. 276]{ewens_2004}. The emphasis here is on predicting how the course of allele frequencies over time depends on their selective effect, genetic drift and linkage. However, the resulting picture of evolution does not include even basic elements of the ecological underpinnings of selection, including dynamic population size and density-dependence.

By contrast, models of phenotypic trait evolution represent the change in phenotypic abundances over time using absolute fitness functions which describe how those traits affect survival and reproduction in particular ecological scenarios. This approach is powerful enough to model eco-evolutionary feedbacks between co-evolving traits, but is generally problem-specific and restricted to only a few traits at a time. These models emphasize invasion from low frequencies and co-existence over frequency trajectories over time. For instance, the adaptive dynamics approach revolves around the analysis of ``invasion fitness'' \citep{diekmann_2004}.

Far less work has been done to generalize beyond particular traits or ecological scenarios to models of fitness that still capture key distinctions between different forms of selection. Perhaps this is not surprising given that fitness is such a complex quantity, dependent on all of a phenotype's functional traits \citep{violle_2007} as well as its biotic and abiotic environment. In most cases, a detailed, trait-based, predictive model of fitness would be enormously complicated and have narrow applicability. It is therefore easy to doubt the feasibility of a simplified, general mathematical treatment of fitness \citep[p. 276]{ewens_2004}. For example, MacArthur's famous $r$/$K$ scheme \citep{macarthur_1962,macarthur_1967} is now almost exclusively known as a framework for understanding life-history traits, and judged on its failure in that role \citep{pianka_1970,stearns_1977,boyce_1984,reznick_2002}. The r/K scheme's original purpose was as an extension of the existing population-genetic treatment of selection to account for population density \citep{macarthur_1962}, but few attempts have been made to develop it further as a mathematical analysis of the major different forms of selection.

Empirical studies of organismal traits have suggested the existence of a few ``primary strategies'' which reflect how different species have responded to selection \citep{winemiller_2015}. The most famous example is Grime's triangle \citep{grime_1974,grime_1977,grime_1988,westoby_1998}. Grime considered two broad determinants of population density: stress (persistent hardship e.g. due to resource scarcity, unfavorable temperatures or toxins) and disturbance (intermittent destruction of vegetation e.g. due to trampling, herbivory, pathogens, extreme weather or fire).  The extremes of these two factors define three primary strategies denoted by C/S/R respectively (Fig. \ref{fig:grimeschematic}): competitors ``C'' excel in low stress, low disturbance environments; stress tolerators ``S'' excel in high stress, low disturbance environments; and ruderals ``R''  excel in low stress, high disturbance environments. Survival is not possible in high-stress, high-disturbance environments. Grime showed that measures of C, S and R across a wide range of plant species are anti-correlated, so that strong C-strategists are weak S and R strategists, and so on, creating a triangular C/S/R ternary plot \citep{grime_1974}. Similar schemes were proposed for insects \citep{southwood_1977}, fishes \citep{winemiller_1992}, and zooplankton \citep{allan_76}. More recently, modern hierarchical clustering techniques have revealed distinct trait clusters in corals analogous to Grime's primary strategies \citep{darling_2012}. Trait classification schemes suggest that, to a first approximation, functional traits contribute to fitness via a few key factors such as intrinsic reproductive rate or stress-tolerance.

@@

Here we explore the interplay between some of these ``key factors'' of fitness in a simplified, territorial model of growth, dispersal and competition, with the broader aim of adding some ecological realism to the time-depedendent, genetically-focused, population genetic models of evolution. To achieve this, we develop a generalized version of the classic lottery model of \cite{chesson_1981}, which has two features that make it well suited for this role.

First, the lottery representation of competition is much simpler than it is in other commonly used ecological models, such as coefficients for the pairwise effects of types on each other (e.g. the $\alpha$ coefficients in the generalized Lotka-Volterra equations), or  explicit resource consumption \citep{tilman_1982}. Mature individuals (``adults'') each require their own territory, whereas newborn individuals (``propagules'') disperse to, and subsequently compete for, territories made available by the death of adults. Territorial contest among propagules leaves a single victorious adult per territory, the victor chosen at random from the propagules present, with probabilities weighted by a coefficient for each type representing competitive ability, akin to a lottery \citep{sale_77}. Second, as we will see, the lottery model is closely connected to one of the central models of population genetics, the Wright-Fisher model of genetic drift.

However, the classic lottery model breaks down at low densities (few propagules dispersing to each territory; see ``Model''). This was not a limitation in its original application to reef fishes, where a huge number of larvae from each species compete to secure territories \citep{chesson_1981}, but is a critical limitation for studying density-dependent selection. We analytically extend the classic lottery model to correctly account for low density behavior.

We broadly follow the spirit of MacArthur's $r$/$K$ selection scheme in that our model is intended to account for fundamentally different forms of selection without getting  entangled in the intricacies of particular ecological scenarios.

However, rather than building directly on MacArthur's formalism and its later extensions using Lotka-Volterra equations to incorporate competition (``$\alpha$-selection'') \citep{gill_1974,case_1974,joshi_2001},

 our model is devised more with Grime's C/S/R scheme in mind, and represents a quantitative formalization of how C/S/R manifests at the level of within-population genotypic evolution (as opposed to phenotypic divergence between species).

 This choice is motivated in part by the substantial empirical support for C/S/R-like schemes, and in part by the failings of the $r$/$K$ low/high density dichotomy --- many growth ability traits will confer advantages at both low and high densities (more details in the Discussion).

Building on the r/K and C/S/R schemes, a central question is how fitness depends on population density, intrinsic birth/death rates and competitive ability.

 However, while trait classification schemes are  grounded in trait data, they are verbal and descriptive rather than mathematical, a recognized hindrance to their broader applicability (e.g. \citealt{tilman_2007}).

In the section ``Model'', we introduce the basic assumptions of our generalized lottery model. Analytical expressions for the change in genotype abundances over time are introduced in section ``Mean field approximation'', with mathematical details relegated to the Appendices. In the following two sections, we then discuss the behavior of rare mutants (invasion and coexistence) and our formalization of Grime's triangle.

@@

The number of individuals of the $i$'th genotype landing in any particular territory is denoted $x_i$. Random dispersal implies that in the limit $T\rightarrow \infty$, with $n_i/T$ held fixed, $x_i$ is Poisson distributed with mean territorial propagule density $l_i=m_i/U$ (this dispersal Poisson distribution is denoted $p_i(x_i)=l_i^{x_i} e^{-l_i}/x_i!$). Although $T$ is finite in our model, we assume that $T$ and the $n_i$ are large enough that $x_i$ is Poisson-distributed to a good approximation (details in Appendix A). Note that the large $n_i$, large $T$ approximation places no restrictions on our densities $n_i/T$, but it does preclude consideration of demographic stochasticity when $n_i$ itself is very small.

@@

So we have only considered the low-frequency invasion behavior of Eq. \eqref{eq:master}. Here we start exploring its full time-dependent behaviour. The broader context for this section is the recurring difficulties in population genetic inference caused by the confounding of population demography and selection --- selective sweeps generate similar genealogies to population bottlenecks \citep{barton_1998}. Current approaches attempt to treat selection and demography separately, typically fitting a demographic model to data on ostensibly neutral sites and then inferring selection against this demographic background, but with mixed success due to linkage \citep{schrider_2016}. Fitness models like Eq. \eqref{eq:master} may help to address to these difficulties since they do not artificially separate the dynamics of abundance and frequency.

>>>>

Evolutionary models differ greatly in their treatment of fitness. In models of genetic evolution, genotypes are typically assigned constant (or frequency-dependent) selection coefficients describing the change in their relative frequencies over time due to differences in viability. This considerably simplifies the mathematics of selection, facilitating greater genetic realism, and can be justified over sufficiently short time intervals \citep[p. 276]{ewens_2004}. The emphasis here is on predicting how the course of allele frequencies over time depends on their selective effect, genetic drift and linkage, and inferring past selection, drift and linkage given a sample of nucleotide sequences. The resulting picture of evolution does not include even basic elements of the ecological underpinnings of selection, such as density-dependence.

By contrast, models of phenotypic trait evolution use absolute fitness functions which describe how some traits of interest affect survival and reproduction in particular ecological scenarios. This approach is powerful enough to model eco-evolutionary feedbacks between co-evolving traits, but is generally problem-specific and restricted to only a few traits at a time. The emphasis here is on invasion from low frequencies and co-existence, rather than frequency trajectories over time. For instance, adaptive dynamics revolves around the analysis of ``invasion fitness'' \citep{diekmann_2004}.

Less work has been done to generalize beyond particular traits or ecological scenarios to models of fitness that still capture key distinctions between different forms of selection. Perhaps this is not surprising given that fitness is such a complex quantity, dependent on all of a phenotype's functional traits \citep{violle_2007} as well as its biotic and abiotic environment. In most cases, a detailed, trait-based, predictive model of fitness would be enormously complicated and have narrow applicability. It is therefore easy to doubt the feasibility of a simplified, general mathematical treatment of fitness \citep[p. 276]{ewens_2004}. For example, MacArthur's famous $r$/$K$ scheme \citep{macarthur_1962,macarthur_1967} is now almost exclusively known as a framework for understanding life-history traits, and judged on its failure in that role \citep{pianka_1970,stearns_1977,boyce_1984,reznick_2002}. The r/K scheme's original purpose was to extend the existing population-genetic treatment of selection to account for population density \citep{macarthur_1962}, but few attempts have been made to develop it further as a mathematical analysis of the major different forms of selection.

On the other hand, empirical studies of organismal traits have suggested the existence of a few ``primary strategies'', reflecting broadly distinct responses to selection \citep{winemiller_2015}. The most famous example is Grime's triangle \citep{grime_1974,grime_1977,grime_1988,westoby_1998}. Grime considered two broad determinants of population density: stress (persistent hardship e.g. due to resource scarcity, unfavorable temperatures or toxins) and disturbance (intermittent destruction of vegetation e.g. due to trampling, herbivory, pathogens, extreme weather or fire).  The extremes of these two factors define three primary strategies denoted by C/S/R respectively (Fig. \ref{fig:grimeschematic}): competitors ``C'' excel in low stress, low disturbance environments; stress tolerators ``S'' excel in high stress, low disturbance environments; and ruderals ``R''  excel in low stress, high disturbance environments. Survival is not possible in high-stress, high-disturbance environments. Grime showed that measures of C, S and R across a wide range of plant species are anti-correlated, so that strong C-strategists are weak S and R strategists, and so on, creating a triangular C/S/R ternary plot \citep{grime_1974}. Similar schemes were proposed for insects \citep{southwood_1977}, fishes \citep{winemiller_1992}, and zooplankton \citep{allan_76}. More recently, modern hierarchical clustering techniques have revealed distinct trait clusters in corals analogous to Grime's primary strategies \citep{darling_2012}. These empirical findings suggest that, to a first approximation, functional traits contribute to fitness via a few key factors such as intrinsic reproductive rate or stress-tolerance.

@@

Here we explore the interplay between some of these ``key factors'' of fitness in a simplified, territorial model of growth, dispersal and competition, with the broader aim of adding some ecological realism to population genetics' time-depedendent, genetically-focused view of evolution. To achieve this, we revisit the classic lottery model of \cite{chesson_1981}, which has two features that make it well suited for this role, and one critical flaw.

The first feature is that the lottery representation of competition is  particularly concise. Mature individuals (``adults'') each require their own territory, whereas newborn individuals (``propagules'') disperse to, and subsequently compete for, territories made available by the death of adults. Territorial contest among propagules leaves a single victorious adult per territory, the victor chosen at random from the propagules present, with probabilities weighted by a coefficient for each type representing competitive ability, akin to a lottery \citep{sale_77}. By comparison, coefficients for the pairwise effects of types on each other (e.g. the $\alpha$ coefficients in the generalized Lotka-Volterra equations and the associated concept of ``$\alpha$-selection''; \citealt{gill_1974,case_1974,joshi_2001}), or  explicit resource consumption \citep{tilman_1982}, are much more complicated. The second feature, as we will see, is the close connection between the lottery model and one of the foundational models of population genetics, the Wright-Fisher model of genetic drift.

The critical flaw is that the classic lottery model breaks down at low densities (few propagules dispersing to each territory). This was not a limitation in its original application to reef fishes \citep{chesson_1981}, which have an enormous number of larvae, but is a critical limitation for studying density-dependent selection. Thus, our first task is to analytically extend the classic lottery model to correctly account for low density behavior (sections ``Model'' and ``Mean field approximation'').

Using our extended lottery model, we then revisit Grime's C/S/R scheme,  and evaluate how C/S/R manifests at the level of within-population genotypic evolution (as opposed to phenotypic divergence between species; sections ``Invasion of rare genotypes and coexistence'' and ``Primary strategies and Grime's triangle''). This represents a ``sanity check'' on our density-dependent lottery model. The resulting formulation of the C/S/R scheme is mathematical, in contrast to Grime's original verbal and descriptive approach, which is a recognized hindrance to the evaluation or broader application of the C/S/R scheme (e.g. \citealt{tilman_2007}).

We then explore some time-dependent behavior of our extended lottery model. The broader context for this section is the recurring difficulties in population genetic inference caused by the similarity of genealogies generated by selective sweeps and population bottlenecks \citep{barton_1998}.

Current approaches attempt to treat selection and demography separately, typically fitting a demographic model to data on ostensibly neutral sites and then inferring selection against this demographic background, but with mixed success due to linkage \citep{schrider_2016}. Fitness models like Eq. \eqref{eq:master} may help to address to these difficulties since they do not artificially separate the dynamics of abundance and frequency. Taking an example inspired by recent studies of rapid, seasonal evolution in \textit{Drosophila} \citep{bergland_14}, we discuss how environmental fluctuations might stabilize polymorphisms in the presence of cyclical population density.

@@

The number of individuals of the $i$'th genotype landing in any particular territory is denoted $x_i$. Random dispersal implies that in the limit $T\rightarrow \infty$, with $n_i$ adjusted to keep $n_i/T$ constant as the limit is taken, $x_i$ is Poisson distributed with mean territorial propagule density $l_i=m_i/U$ (this dispersal Poisson distribution is denoted $p_i(x_i)=l_i^{x_i} e^{-l_i}/x_i!$). Although $T$ is finite in our model, we assume that $T$ and the $n_i$ are large enough that $x_i$ is Poisson-distributed to a good approximation (details in Appendix A). Note that the large $n_i$, large $T$ approximation places no restrictions on our densities $n_i/T$, but it does preclude consideration of demographic stochasticity when $n_i$ itself is very small.

@@

So we have only considered the low-frequency invasion behavior of Eq. \eqref{eq:master}. Here we start exploring its full time-dependent behaviour.
